Bayesian Evidence Synthesis:Opioid Crisis

### 2020-05-01 latex/researchpaper.tex
finally done, my first science research project

%opening
\title{Bayesian Evidence Synthesis:Opioid Crisis}
\author{Hyeongcheol Park*}

@@

Since the total number of overdoses is unknown, this number needs to be estimated. Bayesian statistics can be a way to approach the problem and give us a good estimate of the number with a defensible range of uncertainty. Bayesian statistics is a theory in the field of statistics based on the Bayesian interpretation of probability where probability expresses a degree of belief in an event. \cite{wiki_bayes} Further explanation is provided at the following section.\\

@@

\normalsize 
The number of overdoses is our ultimate target of estimation. To achieve the goal in context of Bayesian statistics, we use our prior belief and available data set about the target of estimation. The posterior belief is coming from both of these sources of the information. Let $\theta$ denote all the unknown variables and let $Y$ denote all the related data sets. The following equation express the idea of Bayesian generative models in a mathmatical form:
\begin{equation}
\label{generic_bayes_thm}
\left.\begin{aligned}
p(\theta| Y ) = \frac{ p(\theta, Y)   }{ p(Y)} = \frac{ p(\theta) p(Y|\theta)}{ p(Y)} \propto p(\theta) p(Y| \theta), \end{aligned}\right.
\end{equation}

where $p(\cdot)$ is probability density (or mass) function. The equation \ref{generic_bayes_thm} states that the posterior belief of unobserved variables, $p(\theta|Y)$, is proportional to the mixture of the prior belief of the variables, $p(\theta)$, and the related data sets given the prior belief, $p(Y|\theta)$. Notice that the vector $\theta$ from equation \ref{generic_bayes_thm} includes parameters and latent variables, and hence it contains the total number of overdoses. \\

Let us assume that we are interested in the opioid crisis from month 1 to month T.  Let the subscript 1:T for any variable denote the variable for all target months and subscript t for any variable denote the variable of the month t. Then the target variable, the total number of overdoses from month 1 to month T  in British Columbia, is denoted by $O_{1:T}$.
\iffalse
Let $O_t$ denote the total number of overdoses in a given month $t$. 
\fi
In Bayesian concepts, the variable of interest $O_{1:T}$ is a vector of fixed numbers but the numbers are yielded from a certain distribution where the distribution describes knowledge about $O_{1:T}$; that is, each $O_t$ is considered to be a sample of a random variable. 
The simplest conceptual model is to take an underlying log-rate $z_t$ for each month $t$ where $z_t$ is independent and identically distributed across months according to a normal distribution with mean $\mu$ and variance $\sigma^2$. \cite{Irvine:modelling} Then let $\lambda_{t}^{OD}$ denote the rate of overdose at month $t$. It is assumed that the number of overdoses in the month, $O_t$, follows Poisson distribution with parameter $\lambda_{t}^{OD}N$ where the size of the population in the region of interest is $N$. Note that the population size, N, is unknown and vary over time in reality but it is assumed known and constant for the investigator here for simplicity. The mechanism is named as the overdose model  and is illustrated in mathmatical form below:

@@

Since estimation of $O_t$ is not straightforward as some of the variables ($\mu$, $\sigma$) determining $O_t$ is not known, we assume there was a survey conducted to help to support the inference about  $O_t$. The survey asked witnesses of opioid overdoses whether someone called an ambulance, for each month. This provides an estimate of the proportion of ambulance call $p_{A,t}$ among the overdoses. For the moment $p_{A,t}$ is assumed constant across time for simplicity. Let $n_{A,t}$ be the sample size of the survey at month $t$, and $x_{A,t}$ be the total number who confirmed that an ambulance was called at the month $t$. It is assumed that $x_{A,t}$ follows a Binomial distribution with number of trials, $n_{A,t}$, and  success probability for each trial, $p_{A,t}$. The mechanism is named as the ambulance call-outs model and is illustrated in mathmatical form below:

@@

Now the estimated $p_{A,t}$ can help to infer $O_t$ if there is a data regarding the number of  ambulance-attended overdoses. Let \(U_t\) be at a time point $t$. In general, the data of ambulance-attended overdoses \(U_t\) can be obtained. It is assumed that  \(U_t\) follows a Binomial distribution: 
\begin{equation}
\label{over_amb}
\left.
U_t \sim Bin(O_t, p_{A,t}).
\right.
\end{equation}

Note that $O_t$ can be estimated because $p_{A,t}$ can be infered from the survey data and the data regarding $U_t$ is given. Figure \ref{simple_draw} shows an example of how the estimation can proceed. The variables that can be observed (N, $U_{1:T}$, and $x_{A,1:T}$) are drawn in squares and the parameters ($\mu, \sigma^2$) and the latent variables ($\lambda_{1:T}, O_{1:T}, p_{A,1:T} $) that are not observarble from possible data sets are drawn in circles. Variables that need prior distributions are indicated with asterisks. The arrows indicate which variables affects which variables; for example, there is an arrow from $(\mu, \sigma^2)$ to $\lambda_{1:T}$ since $\mu, \sigma^2$ are the parameters of $\lambda_{1:T}$, and the two arrows from  $\lambda_{1:T}$ and $N$ to $O_{1:T}$ indicate that both $N$ and $\lambda_{1:T}$ affects the value of $O_{1:T}$ since the product of $\lambda_{1:T}$ and $N$ is the parameter of the distribution of $O_{1:T}$.\\

@@

\begin{equation}
\label{gen_model}
\left.\begin{aligned}
p(\theta, Y) & =p(\mu,\sigma^2,\lambda_t, N, O_t, p_{(A,t)}, n_{(A,t)}, x_{(A,t)}, U_t) \\
& = p(\mu)p(\sigma^2|\mu)p(\lambda_t|\mu, \sigma^2)p(N)p(O_t|\lambda_t, N)p(p_{(A,t)})p(n_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}) \\
& \propto p(\mu)p(\sigma^2)p(\lambda_t|\mu, \sigma^2)p(O_t|\lambda_t, N)p(p_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}),
 \end{aligned}\right.
\end{equation}
where $\theta = (\mu, ,\sigma^2,\lambda_t, O_t, p_{(A,t)})$ and $ Y = ( N, n_{(A,t)}, x_{(A,t)}, U_t)$. The chain rule for random variables \cite{chain_rule} is used to the joint distribution of all variables at the first line of equation \ref{gen_model} such that the second line of the equation is derived.
A priori independence assumptions and conditional independence assumptions were presumed so that the second line of the equation is simplified. For example, it is asumed that  $\sigma^2$ and $\mu$ are independent to each other and $p_{A,t}$ and also all the variables from overdose model \ref{overdose}, ($\mu, \sigma^2, \lambda_t, N, O_t$), are independent to each other. 
Note that as a function of unknown variables, the function from the second line of the equation \ref{gen_model} is proportional to the function from the third line of the equation since $N$ and $n_{(A,t)}$ are presumed to be known and constants such that the density functions of the known variables are merely constants. Therefore the third line of the equation is the ultimate generative model of interest.\\

Conceptunally, generative model yields the posterior distribution $p(\theta|Y)$ on auto pilot since there exists only one distribution of $\theta$ given $Y$. we can use the distribution to make any inference claims that we need to make about parameters or latent variables such as point estimation and credibal interval.\\

@@

Markov Chain Monte Carlo (MCMC) is a detour to obtain posterior samples from $p(\theta|Y)$ without explicitly knowing the distribution.  A Markov chain is a stochastic model describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. In continuous-time, it is known as a Markov process. \cite{markov} Monte Carlo refers to the way how the inference of the target distribution is proceeded. Assessing the properties of a target distribution by generating representative random values is a case of a Monte Carlo simulation and any simulation that samples a lot of random values from a distribution is called a Monte Carlo simulation. \cite{gill}. MCMC technique allows us to generate Monte Carlo realizations of samples $\theta_{(i)}$ which are dependent, with the distribution of each sample $p(\theta_{(i)})$ converging to $p(\theta|Y)$.\cite{paul}.

@@

The No-U-Turn Sampler (NUTS) is a Hamiltonian Monte Carlo (HMC) Method that is used in pyMC3. HMC is a MCMC algorithm that avoids the random walk behavior and sensitivity to correlated parameters that plague many MCMC methods by taking a series of steps informed by first-order gradient information. These features allow it to converge to high-dimensional target distributions much more quickly than simpler methods such as random walk Metropolis or Gibbs sampling. However HMC’s performance is highly sensitive to two user-specified parameters: a step size and a desired number of steps. \cite{hoffman2011nouturn} Moreover, NUTS struggles to samples discrete random variables and hence it is recommended to approximate discrete random variables to continuous ones. The Poisson distribution from \ref{overdose} was approximated to Gamma distribution with the same parameter ($\lambda_{t}^{OD}N$) in pyMC3.

@@

First, it is mandatory to throw away some initial samples of Markov Chain realizations. Note that a value of $\theta_{(i)}$  from a lot of iteration better represents a sample from the posterior distribuion than some $\theta_{(j)}$ value from some early period of iterations ($i > j$). Since the convergence takes enough number of iterations, it is necessary to discard a chunk of the iterations from $\theta_{(1)}$ to $\theta_{(n)}$ for a large number $n$ enough to obtain the proper samples to represent the samples of interest from the posterior distribution. The period of iterations to be discarded is called burn-in period.
\\

Secondly, it is strongly recommended to have multiple starting points to assure the convergence. A visual way to check the convergence is to see a plot of  iteration numbers in x axis and their values for each iteration in y axis (trace plot). Note that $p(\theta_{(i)})$ converges to  $p(\theta|Y)$ regardless of the initial position of $\theta_{(1)}$, but the location of a initial position affects the number of iterations of MCMC algorithm for the convergence. Looking at only a single chain of iterations and their values may mislead us to believe that the number of iterations is enough while the trace of the values are staying at a local area before it reaches to the target distribution. By having more than one initial points and checking that the chains get mixed well and are stationary at a certain iteration point, we can verify that the algorithm reaches the desirable target distribution. 
\\

Lastly, numerical measures such as $\hat{R}$ and effective sample size can help us to examine the degree of convergence as well as looking at the trace plot of multiple chains of iterations. $\hat{R}$ is a measure of convergence where values for $\hat{R}$ near 1 suggest convergence. Effective sample size is a crude measure of the number of independent (effective) sample size from the total number of samples generated from a certain Markov chain. Both criteria is accessible after a generative model is fitted from a summary report of MCMC packages. The values of $\hat{R}$ and effective sample size in this report are obtainaed from JAGS. \\

@@

The achievement of the simulation was done by Markov chain Monte Carlo (MCMC). All examples here were performed in Python 3.8 with the library pyMC3 \cite{pymc3} and also in R 3.6.2 \cite{r_lan} with the JAGS 4.3.0 \cite{jags} library. All visualizations were the results from Python and pyMC3. Fitting was performed on a GHz Intel Core i5 with 8GM of LPDD3 RAM and typically had wall times under ten minutes. Code for all examples in this study are provided.

@@

Figure \ref{ppc_ut} is the boxplot of posterior predictive samples of $U_t$. It is shown that the posterior predictive estimates of $U_t$ are failry accurate, for the same two reasons regarding the accuracy of the posterior distribution of $O_t$. It is more obvious here that the range of the boxplots from higher $O_t$ values (t=1, 7, 11) is wider than the other ranges of the boxplots from smaller estimates of $O_t$ (all t values but 1,7, 11). Notice that the relative ranges of figure \ref{ppc_ut} follow the ones of figure \ref{pst_ot}; for months where $U_t$ values are higher, the values of $O_t$ are also higher than the average (t=1, 7, 11).\\

@@

Figure \ref{contam_xt}  gives the boxplots of posterior predictive samples of $x_{A,t}$ where $p_{A,t}$ is underestimated and overestimated respectively from the survey data. It is seen that none of the contaminations of $p_{A,t}$ leads an effect of contamination on $x_{A,t}$; the actual simulated points (red dots) are close to the medians from the two boxplots. However, notice that the range of the estimated values are different between the two boxplots; the median from Figure \ref{under_xa} is around 600 whereas the median from Figure \ref{over_xa} is around 900. This implies that the posterior predictive samples match the biased given data sets (with $p_{A,t}=0.6$ and $p_{A,t}=0.9$ respectively) but the samples cannot estimate the unbiased true value of $p_{A,t}$, which is 0.8. This is because overdose model (\ref{overdose}) does not affect the result from the ambulance model (\ref{ambulance}); only the ambulance model has an effect on the overdose model as it is seen from figure \ref{simple_draw}.

@@

\subsection{Result: Unobservable $N$}
In practice, it is reasonable to assume that investigators only have a general idea about the total population size of the reigion rather that they know the exact value of the total population size. The assumption of the known value of the total population size for a region, $N$, from the previous section changes to be unknown in this section to see the impact of the likelihood over the posterior distributions of variables of interest. The generative model at a month $t$ can be written as follows given some a priori assumptions:

\begin{equation}
\label{gen_model_unk_N}
\left.\begin{aligned}
p(\theta, Y) & =p(\mu,\sigma^2,\lambda_t, N, O_t, p_{(A,t)}, n_{(A,t)}, x_{(A,t)}, U_t) \\
& = p(\mu)p(\sigma^2|\mu)p(\lambda_t|\mu, \sigma^2)p(N)p(O_t|\lambda_t, N)p(p_{(A,t)})p(n_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}) \\
& \propto p(\mu)p(\sigma^2)p(\lambda_t|\mu, \sigma^2)p(N)p(O_t|\lambda_t, N)p(p_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}),
\end{aligned}\right.
\end{equation}

where $\theta = (\mu, ,\sigma^2,\lambda_t, O_t, p_{(A,t)}, N)$ and $ Y = (n_{(A,t)}, x_{(A,t)}, U_t)$. Notice that $N$ is an element of $\theta$ instead of an element of $Y$ and the model contains a prior distribution of $N$. The chain rule for random variables \cite{chain_rule} is used to the joint distribution of all variables at the first line of equation \ref{gen_model_unk_N} such that the second line of the equation is derived.
A priori independence assumptions and conditional independence assumptions were presumed so that the second line of the equation is simplified. The model is shown graphically in figure \ref{draw_N} where $N$ is a latent variable with a prior distribution needed.\\

@@

\begin{equation}
\label{in_prior_N}
N \sim Normal(10000,1000)
\text{ (informative prior for the total population size),}
\end{equation}

where the prior information suggests that $N$ is most likely around 10000 and $\pm 20\%$ of the point estimation covers  95\% of the symmetric support of $N$.
\subsubsection{Posterior Distribution}
Both MCMC algorithms from pyMC3 and JAGS suffered from relatievely long running time of the algorithms and the convergence issue of the trace of $N$. The reason of having the stationarity issue is that too much inference has been asked to the data sets. Only two data sets, $U_{1:T}$ and $(n_{A,1:T}, x_{A,1:T})$, are given and the task was to generate posterior samples of 5 random variables, $N, \mu, \sigma^2, \lambda_{1:T}, O_{1:T}, p_{A,1:T}$. Figure \ref{trace_N} shows the trace plot and the posterior density plot of $N$. From the trace plot of $N$ it is seen that the chains are mixed well but not stationary and thus not convenging; two chains of the variable covered a common distribution but neither sequence appeared stable even after 10000000 iterations. The effective sample size for N was only 56271. The posterior distribution seems to be roughly symmetric at the center 10000. 
\begin{figure}[htb]
	\centering
	\includegraphics[width=1\linewidth]{Figures/not_mixed_N.png}
	\caption[Unobserved $N$: trace plot of latent variable $N$]{Trace plot of latent variable $N$}
	\label{trace_N}
\end{figure}

@@

Figure \ref{ppc_unk_n} is the boxplot of posterior predictive samples of $U_t$ and $x_{A,t}$. It is shown that the posterior predictive estimates of $U_t$ and $x_{A,t}$ are failry accurate for the same reason regarding the accuracy of the posterior distribution of $O_t$.

@@

Bayesian statistics and generative model is an outstanding approach for statistical inference which incorporate available data sets and our  prior knowlede and information. Bayesian model can cope with prior research outcomes as our prior information whereas frequentist approach cannot absorb the knowledge in their models. Furthermore, Bayesian modeling is intuitive to us since Bayesian viewpoint to probability corresponds to our conceptual perspective about probability. \\

However, Bayesian models have their own weaknesses. In cases where the problems are immensely complicated, the models suffer much more than the Frequentist models from long computational time of MCMC algorithms. This issue has been being relived as the general computing power of machines has been improved and decent MCMC algorithms have been developed to reduce waiting time although patient waiting is still a virtue to Bayesian statisticians.\\

Note that the robustness of models shoud be confirmed. If there exist some contaminations of data sets, the result of posterior estimations may not be trustworthy and therefore it is needed to confirm how robust the result of models is from the contaimiations. Likewise, it is generally recommended to apply more than one MCMC algorithms to the generative model to confirm that Monte Carlo realizations of  samples are robust to different types of the algorithms.\\

Lastly, notice that it is not reasonable to provide few data sets and ask Bayesian generative models to yield decent results; the models are not a god-like solver for every qnatitative problems. It is shown from the previous section that asking about the total population size $N$ is too much to ask for the model while the survey data $(n_{A,1:T}, x_{A,1:T})$ and the data of the number of ambulance-attended overdoses $(U_{1:T})$ are only provided.	\\

>>>>

%opening
\title{Bayesian Evidence Synthesis:Opioid Crisis}
\author{Hyeongcheol Park}

@@

Since the total number of overdoses is unknown, this number needs to be estimated. Bayesian statistics can be a way to approach the problem and give us a good estimate of the number with a defensible range of uncertainty. Bayesian statistics is a theory in the field of statistics based on the Bayesian interpretation of probability where probability expresses a degree of belief in an event. \cite{wiki_bayes} Further explanation is provided in the following section.\\

@@

\normalsize 
The number of overdoses is our ultimate target of estimation. To achieve the goal in context of Bayesian statistics, we use our prior belief and available data set about the target of estimation. The posterior belief is coming from both of these sources of information. Let $\theta$ denote all the unknown variables and let $Y$ denote all the related data sets. The following equation expresses the idea of Bayesian generative models in a mathmatical form:
\begin{equation}
\label{generic_bayes_thm}
\left.\begin{aligned}
p(\theta| Y ) = \frac{ p(\theta, Y)   }{ p(Y)} = \frac{ p(\theta) p(Y|\theta)}{ p(Y)} \propto p(\theta) p(Y| \theta), \end{aligned}\right.
\end{equation}
where $p(\cdot)$ is probability density (or mass) function. The equation \ref{generic_bayes_thm} states that the posterior belief of unobserved variables, $p(\theta|Y)$, is proportional to the mixture of the prior belief of the variables, $p(\theta)$, and the related data sets given the prior belief, $p(Y|\theta)$. Notice that the vector $\theta$ from equation \ref{generic_bayes_thm} includes parameters and latent variables, and hence it contains the total number of overdoses. \\

Let us assume that we are interested in the opioid crisis from month 1 to month T.  Let the subscript 1:T for any variable denote the variable for all target months and subscript t for any variable denote the variable of the month t. Then the target variable, the total number of overdoses from month 1 to month T  in British Columbia, is denoted by $O_{1:T}$.
\iffalse
Let $O_t$ denote the total number of overdoses in a given month $t$. 
\fi
Applying Bayesian concepts, the variable of interest $O_{1:T}$ is a vector of fixed numbers but the numbers are yielded from a certain distribution where the distribution describes knowledge about $O_{1:T}$; that is, each $O_t$ is considered to be a sample of a random variable. 
The simplest conceptual model is to take an underlying log-rate $z_t$ for each month $t$, where $z_t$ is independent and identically distributed across months according to a normal distribution with mean $\mu$ and variance $\sigma^2$. \cite{Irvine:modelling} Then let $\lambda_{t}^{OD}$ denote the rate of overdose at month $t$. It is assumed that the number of overdoses in the month, $O_t$, follows a Poisson distribution with parameter $\lambda_{t}^{OD}N$ where the size of the population in the region of interest is $N$. Note that the population size, N, is unknown and vary over time in reality but it is assumed known and constant for the investigator here for simplicity. The mechanism is named as the overdose model  and is illustrated in mathmatical form below:

@@

Since estimation of $O_t$ is not straightforward as some of the variables ($\mu$, $\sigma$) determining $O_t$ are not known, we assume there was a survey conducted to help to support the inference about  $O_t$. The survey asked witnesses of opioid overdoses whether someone called an ambulance, for each month. This provides an estimate of the proportion of ambulance call $p_{A,t}$ among the overdoses. For the moment $p_{A,t}$ is assumed constant across time for simplicity. Let $n_{A,t}$ be the sample size of the survey at month $t$, and $x_{A,t}$ be the total number who confirmed that an ambulance was called at the month $t$. It is assumed that $x_{A,t}$ follows a Binomial distribution with number of trials, $n_{A,t}$, and  success probability for each trial, $p_{A,t}$. The mechanism is named as the ambulance call-outs model and is illustrated in mathmatical form below:

@@

Now the estimated $p_{A,t}$ can help to infer $O_t$ if there is data regarding the number of  ambulance-attended overdoses. Let \(U_t\) be at a time point $t$. In general, the data of ambulance-attended overdoses \(U_t\) can be obtained. It is assumed that  \(U_t\) follows a Binomial distribution: 
\begin{equation}
\label{over_amb}
\left.
U_t \sim Bin(O_t, p_{A,t}).
\right.
\end{equation}

Note that $O_t$ can be estimated because $p_{A,t}$ can be infered from the survey data and the data regarding $U_t$ is given. Figure \ref{simple_draw} shows an example of how the estimation can proceed. The variables that can be observed (N, $U_{1:T}$, and $x_{A,1:T}$) are drawn in squares and the parameters ($\mu, \sigma^2$) and the latent variables ($\lambda_{1:T}, O_{1:T}, p_{A,1:T} $) that are not observable from possible data sets are drawn in circles. Variables that need prior distributions are indicated with asterisks. The arrows indicate which variables affects which variables; for example, there is an arrow from $(\mu, \sigma^2)$ to $\lambda_{1:T}$ since $\mu, \sigma^2$ are the parameters of $\lambda_{1:T}$, and the two arrows from  $\lambda_{1:T}$ and $N$ to $O_{1:T}$ indicate that both $N$ and $\lambda_{1:T}$ affects the value of $O_{1:T}$ since the product of $\lambda_{1:T}$ and $N$ is the parameter of the distribution of $O_{1:T}$.\\

@@

\begin{equation}
\label{gen_model}
\left.\begin{aligned}
p(\theta, Y) & =p(\mu,\sigma^2,\lambda_t, N, O_t, p_{(A,t)}, n_{(A,t)}, x_{(A,t)}, U_t) \\
& = p(\mu)p(\sigma^2|\mu)p(\lambda_t|\mu, \sigma^2)p(N)p(O_t|\lambda_t, N)p(p_{(A,t)})p(n_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}) \\
& \propto p(\mu)p(\sigma^2)p(\lambda_t|\mu, \sigma^2)p(O_t|\lambda_t, N)p(p_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}),
 \end{aligned}\right.
\end{equation}
where $\theta = (\mu, ,\sigma^2,\lambda_t, O_t, p_{(A,t)})$ and $ Y = ( N, n_{(A,t)}, x_{(A,t)}, U_t)$. The chain rule for random variables \cite{chain_rule} is used to the joint distribution of all variables at the first line of equation \ref{gen_model} such that the second line of the equation is derived.
A priori independence assumptions and conditional independence assumptions were presumed so that the second line of the equation is simplified. For example, it is asumed that  $\sigma^2$ and $\mu$ are independent of each other and $p_{A,t}$ and also all the variables from overdose model \ref{overdose}, ($\mu, \sigma^2, \lambda_t, N, O_t$), are independent of each other. 
Note that as a function of unknown variables, the function from the second line of the equation \ref{gen_model} is proportional to the function from the third line of the equation since $N$ and $n_{(A,t)}$ are presumed to be known and constants such that the density functions of the known variables are merely constants. Therefore the third line of the equation is the ultimate generative model of interest.\\

The generative model yields the posterior distribution $p(\theta|Y)$ on auto pilot, since there exists only one distribution of $\theta$ given $Y$. we can use the distribution to make any inference claims that we need to make about parameters or latent variables such as point estimation and credibal interval.\\

@@

Markov Chain Monte Carlo (MCMC) is a detour to obtain posterior samples from $p(\theta|Y)$ without explicitly knowing the distribution.  A Markov chain is a stochastic model describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. In continuous-time, it is known as a Markov process. \cite{markov} Monte Carlo refers to how inference about the target distribution proceeds. Assessing the properties of a target distribution by generating representative random values is a case of a Monte Carlo simulation and any simulation that samples a lot of random values from a distribution is called a Monte Carlo simulation. \cite{gill}. MCMC technique allows us to generate Monte Carlo realizations of samples $\theta_{(i)}$ which are dependent, with the distribution of each sample $p(\theta_{(i)})$ converging to $p(\theta|Y)$.\cite{paul}.

@@

The No-U-Turn Sampler (NUTS) is a Hamiltonian Monte Carlo (HMC) Method that is used in pyMC3. HMC is a MCMC algorithm that avoids the random walk behavior and sensitivity to correlated parameters that plague many MCMC methods by taking a series of steps informed by first-order gradient information. These features allow it to converge to high-dimensional target distributions much more quickly than simpler methods such as random walk Metropolis or Gibbs sampling. However HMC’s performance is highly sensitive to two user-specified parameters: a step size and a desired number of steps. \cite{hoffman2011nouturn} Moreover, NUTS struggles to sample discrete random variables and hence it is recommended to approximate discrete random variables with continuous ones. The Poisson distribution from \ref{overdose} was approximated to Gamma distribution with the same parameter ($\lambda_{t}^{OD}N$) in pyMC3.

@@

First, it is mandatory to throw away some initial samples of Markov Chain realizations. Note that a value of $\theta_{(i)}$  from a lot of iteration better represents a sample from the posterior distribution than some $\theta_{(j)}$ value from some early period of iterations ($i > j$). Since the convergence takes some number of iterations, it is necessary to discard a chunk of the iterations from $\theta_{(1)}$ to $\theta_{(n)}$ for a large number $n$ enough to obtain the proper samples to represent the samples of interest from the posterior distribution. The period of iterations to be discarded is called the burn-in period.
\\

Secondly, it is strongly recommended to have multiple starting points to assure the convergence. A visual way to check the convergence is to see a plot of  iteration numbers in x axis and their values for each iteration in y axis (trace plot). Note that $p(\theta_{(i)})$ converges to  $p(\theta|Y)$ regardless of the initial position of $\theta_{(1)}$, but the location of a initial position affects the number of iterations of MCMC algorithm for the convergence. Looking at only a single chain of iterations and their values may mislead us to believe that the number of iterations is enough, with the trace of the values are staying at a local area before it reaches to the target distribution. By having more than one initial point and checking that the chains mix well and are stationary at a certain iteration point, we can verify that the algorithm reaches the desirable target distribution. 
\\

Lastly, numerical measures such as $\hat{R}$ and effective sample size can help us to examine the degree of convergence as well as looking at the trace plot of multiple chains of iterations. $\hat{R}$ is a measure of convergence where values for $\hat{R}$ near 1 suggest convergence. Effective sample size is a crude measure of the number of independent (effective) sample size from the total number of samples generated from a certain Markov chain. Both criteria are accessible after a generative model is fitted from a summary report of MCMC packages. The values of $\hat{R}$ and effective sample size in this report are obtained from JAGS. \\

@@

The implementation of the simulation was done by Markov chain Monte Carlo (MCMC). All examples here were performed in Python 3.8 with the library pyMC3 \cite{pymc3} and also in R 3.6.2 \cite{r_lan} with the JAGS 4.3.0 \cite{jags} library. All visualizations were the results from Python and pyMC3. Fitting was performed on a GHz Intel Core i5 with 8GM of LPDD3 RAM and typically had wall times under ten minutes. Code for all examples in this study are provided.

@@

Figure \ref{ppc_ut} is the boxplot of posterior predictive samples of $U_t$. It is shown that the posterior predictive estimates of $U_t$ are accurate, for the same two reasons regarding the accuracy of the posterior distribution of $O_t$. It is more obvious here that the range of the boxplots from higher $O_t$ values (t=1, 7, 11) is wider than the other ranges of the boxplots from smaller estimates of $O_t$ (all t values but 1,7, 11). Notice that the relative ranges of figure \ref{ppc_ut} follow the ones of figure \ref{pst_ot}; for months where $U_t$ values are higher, the values of $O_t$ are also higher than the average (t=1, 7, 11).\\

@@

Figure \ref{contam_xt}  gives the boxplots of posterior predictive samples of $x_{A,t}$ where $p_{A,t}$ is underestimated and overestimated respectively from the survey data. It is seen that none of the contaminations of $p_{A,t}$ leads an effect of contamination on $x_{A,t}$; the actual simulated points (red dots) are close to the medians from the two boxplots. However, notice that the range of the estimated values are different between the two boxplots; the median from Figure \ref{under_xa} is around 600 whereas the median from Figure \ref{over_xa} is around 900. This implies that the posterior predictive samples match the biased given data sets (with $p_{A,t}=0.6$ and $p_{A,t}=0.9$ respectively) but the samples cannot estimate the unbiased true value of $p_{A,t}$, which is 0.8. This is because overdose model (\ref{overdose}) does not affect the result from the ambulance model (\ref{ambulance}); only the ambulance model has an effect on the overdose model as is seen from figure \ref{simple_draw}.

@@

\subsection{Result: Unobservable $N$}
In practice, it is reasonable to assume that investigators only have a general idea about the total population size of the region rather that they know the exact value of the total population size. The assumption of the known value of the total population size for a region, $N$, from the previous section changes to be unknown in this section to see the impact of the likelihood over the posterior distributions of variables of interest. The generative model at a month $t$ can be written as follows given some a priori assumptions:

\begin{equation}
\label{gen_model_unk_N}
\left.\begin{aligned}
p(\theta, Y) & =p(\mu,\sigma^2,\lambda_t, N, O_t, p_{(A,t)}, n_{(A,t)}, x_{(A,t)}, U_t) \\
& = p(\mu)p(\sigma^2|\mu)p(\lambda_t|\mu, \sigma^2)p(N)p(O_t|\lambda_t, N)p(p_{(A,t)})p(n_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}) \\
& \propto p(\mu)p(\sigma^2)p(\lambda_t|\mu, \sigma^2)p(N)p(O_t|\lambda_t, N)p(p_{(A,t)})p(x_{(A,t)}|p_{(A,t)},n_{(A,t)})p(U_t|O_t, p_{(A,t)}),
\end{aligned}\right.
\end{equation}
where $\theta = (\mu, ,\sigma^2,\lambda_t, O_t, p_{(A,t)}, N)$ and $ Y = (n_{(A,t)}, x_{(A,t)}, U_t)$. Notice that $N$ is an element of $\theta$ instead of an element of $Y$ and the model contains a prior distribution of $N$. The chain rule for random variables \cite{chain_rule} is used to the joint distribution of all variables at the first line of equation \ref{gen_model_unk_N} such that the second line of the equation is derived.
A priori independence assumptions and conditional independence assumptions were presumed so that the second line of the equation is simplified. The model is shown graphically in figure \ref{draw_N} where $N$ is a latent variable with a prior distribution needed.\\

@@

\begin{equation}
\label{in_prior_N}
N \sim Normal(10000,1000)
\text{ (informative prior for the total population size),}
\end{equation} 
where the prior information suggests that $N$ is most likely around 10000 and $\pm 20\%$ of the point estimation covers  95\% of the symmetric support of $N$.
\subsubsection{Posterior Distribution}
Both MCMC algorithms from pyMC3 and JAGS suffered from relatively long running time of the algorithms and a convergence issue for the trace of $N$. The reason of having the stationarity issue is that too much inference has been asked of the data sets. Only two data sets, $U_{1:T}$ and $(n_{A,1:T}, x_{A,1:T})$, are given and the task was to generate posterior samples of 5 random variables, $N, \mu, \sigma^2, \lambda_{1:T}, O_{1:T}, p_{A,1:T}$. Figure \ref{trace_N} shows the trace plot and the posterior density plot of $N$. From the trace plot of $N$ it is seen that the chains are mixed well but not stationary and thus not convenging; two chains of the variable covered a common distribution but neither sequence appeared stable even after 10000000 iterations. The effective sample size for N was only 56271. The posterior distribution seems to be roughly symmetric at the center 10000. 
\begin{figure}[htb]
	\centering
	\includegraphics[width=1\linewidth]{Figures/not_mixed_N.png}
	\caption[Unobserved $N$: trace plot of latent variable $N$]{Trace plot of latent variable $N$}
	\label{trace_N}
\end{figure}

@@

Figure \ref{ppc_unk_n} is the boxplot of posterior predictive samples of $U_t$ and $x_{A,t}$. It is shown that the posterior predictive estimates of $U_t$ and $x_{A,t}$ are accurate for the same reason regarding the accuracy of the posterior distribution of $O_t$.

@@

Bayesian statistics and generative model is an outstanding approach for statistical inference which incorporate available data sets and our  prior knowledge and information. Bayesian model can cope with prior research outcomes as our prior information whereas frequentist approach cannot absorb such knowledge in their models. Furthermore, Bayesian modeling is intuitive to us since the Bayesian viewpoint to probability corresponds to our conceptual perspective about probability. \\

However, Bayesian models have their own weaknesses. In cases where the problems are immensely complicated, the models suffer much more than the Frequentist models from long computational time of MCMC algorithms. This issue has been being relived as the general computing power of machines has been improved and decent MCMC algorithms have been developed to reduce waiting time, although patient waiting is still a virtue to Bayesian statisticians.\\

Note that the robustness of models shoud be confirmed. If there exist some contaminations of data sets, the result of posterior estimations may not be trustworthy and therefore it is necessary to confirm how robust the results of models are to the contaimiations. Likewise, it is generally recommended to apply more than one MCMC algorithm to the generative model to confirm that Monte Carlo realizations of  samples are robust to different types of the algorithms.\\

Lastly, notice that it is not reasonable to provide few data sets and ask Bayesian generative models to yield decent results; the models are not a god-like solver for every quantitative problem. It is shown in the previous section that asking about the total population size $N$ is too much to ask for the model when only the survey data $(n_{A,1:T}, x_{A,1:T})$ and the data of the number of ambulance-attended overdoses $(U_{1:T})$ are provided.	\\
